---
title: "Lesson 1 - Solutions"
author: "Anna Josephson"
date: "2025-07-16"
output: html_document
---

  ## Setup
  
```{r load-packages}
library(arrow)
library(dplyr)
library(ggplot2)
```

```{r load-data}
# Load the data
merged_data <- read_parquet("merged_data.parquet", stringsAsFactors = TRUE)

# Create state subsets
subset_data_az <- subset(merged_data, STATE_ABBR_2010SVI == "AZ")
subset_data_ca <- subset(merged_data, STATE_ABBR_2010SVI == "CA")
```

---
  
## Activity 1: Explore a Different County
  
We'll use **Yuma County** (AZ) as an example.

```{r yuma-stats}
# Calculate Yuma stats
print(yuma_mean_2010 <- mean(subset_data_az$upward_mobility_rate_2010[subset_data_az$COUNTY_2010SVI == "Yuma County"]))
print(yuma_sd_2010 <- sd(subset_data_az$upward_mobility_rate_2010[subset_data_az$COUNTY_2010SVI == "Yuma County"]))

print(yuma_mean_2020 <- mean(subset_data_az$upward_mobility_rate_2020[subset_data_az$COUNTY_2010SVI == "Yuma County"]))
print(yuma_sd_2020 <- sd(subset_data_az$upward_mobility_rate_2020[subset_data_az$COUNTY_2010SVI == "Yuma County"]))

# Compare to state averages
print(upward_mean_2010_az <- mean(subset_data_az$upward_mobility_rate_2010))
print(upward_mean_2020_az <- mean(subset_data_az$upward_mobility_rate_2020))
```

---

## Activity 2: Boxplots

```{r boxplot-yuma}
# 2010 boxplot
ggplot(subset_data_az[subset_data_az$COUNTY_2010SVI == "Yuma County", ],
       aes(x = "2010", y = upward_mobility_rate_2010)) +
  geom_boxplot() +
  labs(title = "Yuma County Mobility 2010", y = "Mobility Rate")

# 2020 boxplot
ggplot(subset_data_az[subset_data_az$COUNTY_2010SVI == "Yuma County", ],
       aes(x = "2020", y = upward_mobility_rate_2020)) +
  geom_boxplot() +
  labs(title = "Yuma County Mobility 2020", y = "Mobility Rate")
```

```{r save-boxplots, eval=FALSE}
ggsave("output/yuma_mobility_2010.pdf")
ggsave("output/yuma_mobility_2020.pdf")
```

---

## Activity 3: Filter Challenge

```{r filter-high-mobility}
# Filter and count
filtered_high_mobility <- subset_data_az %>%
  filter(upward_mobility_rate_2010 > 1)

# How many rows?
nrow(filtered_high_mobility)

# Which counties?
table(filtered_high_mobility$COUNTY_2010SVI)
```

---

## Activity 4: Reflect and Extend

### Example Research Question
> Is upward mobility higher in counties with more educational access?

### Sample Questions to Explore
- When should I use `filter()` vs `subset()`?
- What does it mean if I get `NaN` in `mean()` or `sd()`?

---

## End of Solution Guide
